---
title: "Milestone Report for NLP Capstone Project"
subtitle: "Data Science Specialization on Coursera"
author: "Vadim K."
date: "22 August 2017"
output: 
    html_document:
        keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "data")

```

## Synopsis

The purpose of this document is to provide a milestone report on developing an 
interactive text prediction web app.  
This developement is a part of the capstone project that 
wraps up Coursera Data Science specialization offered by Johns Hopkins University (Baltimore, Maryland).

This application will predict the next word a user most likely will type 
based on the words he/she already typed.  
The server part of application will run a word prediction algorithm that will be finalized later in the project. 
This algorithm is based on N-gram language model built from [HC Corpora](https://web-beta.archive.org/web/20160930083655/http://www.corpora.heliohost.org/aboutcorpus.html) (a set of files with millions lines of text in different languages collected from publicly available sources by a web crawler).

In the present document I will give a brief overview of the first steps of the project 
that were already achieved such as downloading, cleaning and exploring the data. 
I will also share couple of issues I faced and some interesting findings in the data.  
In conclusion you'll find some info about my plans for creating the prediction algorithm and the expected constraints.


## Data Processing

```{r, include = FALSE, warning = FALSE, message = FALSE}
library(knitr)
library(ggplot2)
library(gridExtra)
library(wordcloud)
```

The R packages used during analysis are


```{}
library(stringr)    # to count words
library(stringi)    # to count max characters (seems faster than stringr)
library(dplyr)      # to rename columns of summary table
library(readtext)   # to read sample file into a Corpus
library(quanteda)   # for all the text mining
library(data.table) # to hold the frequency tables
library(ggplot2)    # frequencies visualization
library(wordcloud)  # word clouds
```


#### _Loading Data_

The data was obtained from the archive available under following [link](https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip).  
It contains texts from 3 types of sources (blogs, news, Twitter) in 4 different locales (en_US, de_DE, ru_RU and fi_FI).  
Files in English were taken for this project (i.e. `"en_US.blogs.txt"; "en_US.news.txt"; "en_US.twitter.txt"`).  


<span style="color:gray">
_There was an issue with loading News file: it containes ASCII `Substitute` character on line 77259, 
this character (decimal 26 or 0x1A) corresponds to Ctrl+Z in Windows so the function `readLines` truncates the file at it and imports only 77259 lines.  
Solution is to launch `readLines` in binary mode (credits to  [this Q on StackOverflow](https://stackoverflow.com/questions/15874619/reading-in-a-text-file-with-a-sub-1a-control-z-character-in-r-on-windows))_
</span>



#### _Summary statistics_
Here is some summary statistics for the chosen files:

```{r}
load('summary.table.RData')
kable(summary.table)
```


****************


#### _Sampling & Cleaning the Data_

As the files are pretty big for the further analysis samples of 100k lines 
were taken from each of them.

I also had to do another cleaning for the sample files, as after the first round of 
exploratory analysis some strange word pairs were discovered among the most popular ones 
(e.g. `"u 009f"`).  
After tracing back the issue many lines with non-ASCII characters were found  

they looked like this in R (being read in binary mode)  
`"Girls night with and ! \xf0\u009f\u0092\u009c"`  

and like this in the original txt file  
`Girls night with and ! 💜`  
<span style="color:gray">_Example from line #1054179 of Twitter file_</span>


## Exploring the Data

For exploring the data I experimented with 3 different packages (`tm`, `tidytext` and `quanteda`).  
Personally I found 'quanteda' the most convenient for this project, so all the further text mining is done with it.  

Also two different approaches were studied of taking the 3 files separately or combining them together. 
The decision was to combine.  

So the following steps were done on text mining stage:  

* The Corpus has been read with `readtext`
* Split into sentences (I found it more logical for further tokenization to treat each sentence separately
not just each line) --> so we got **526 022 sentences**
* Sentences then were split by words (1-gram 'tokens' object) and following transformation done:
    + numbers, punctuation and different symbols removed
    + all words to lower case
    + English stop words removed <span style="color:gray">_(For the sake of exploring and building word clouds. Won't do this for prediction algorithm)_</span>
    + Stemming applied <span style="color:gray">_(btw I find quanteda stemming being the most efficient one)_</span>
* 3 more 'tokens' objects were created (holding 2-, 3- and 4-grams respectively)
* Document-feature matrices created from 'tokens' objects, holding
    +   **112 600 unique unigrams**
    + **2 387 021 unique bigrams**
    + **3 596 041 unique trigrams**
    + **3 295 301 unique four-grams**
* Data.tables were created to hold the frequences of n-grams and be used for further plotting  

And below you can check-out the visualization of most popular n-grams in the studied dataset:

```{r}
load('dt.RData')
plot_n_gram <- function(data, title) {
    ggplot(head(data, 20), aes(x = reorder(ngram, count), y = count)) +
    geom_col() + 
    theme(plot.title = element_text(hjust = 0.5),
          axis.title.x = element_blank()) +
    coord_flip() +
    ggtitle(title) + xlab(NULL)
}

g1 <- plot_n_gram(dt1, "Most frequent unigrams")
g2 <- plot_n_gram(dt2, "Most frequent bigrams")
g3 <- plot_n_gram(dt3, "Most frequent trigrams")
g4 <- plot_n_gram(dt4, "Most frequent four-grams")
```

```{r fig.width = 10, fig.height = 5}
grid.arrange(g1, g2, nrow = 1)
```

```{r fig.width = 10}
grid.arrange(g3, g4, nrow = 1)
```


And here is another representation of unigrams and bigrams frequencies - by Word Clouds:

```{r include = FALSE, warning = FALSE, message = FALSE}
plot_wcloud <- function(data, scale, n, title) {
    layout(matrix(c(1, 2), nrow = 2), heights = c(2, 40))
    par(mar = rep(0, 4))
    plot.new()
    text(x = 0.5, y = 0.3, title)
    
    wordcloud(words = data$ngram,
          freq = data$count, 
          colors = brewer.pal(6, "Dark2"), 
          scale = c(scale, 0.2),
          rot.per = 0.3,
          max.words = n,
          random.order = FALSE)
}
```

```{r warning = FALSE, message = FALSE, fig.height = 6}
plot_wcloud(dt1, 4, 200, "Unigrams Word Cloud")
```
```{r warning = FALSE, message = FALSE, fig.height = 6}
plot_wcloud(dt2, 4, 120, "Bigrams Word Cloud")
```

#### _Couple of interesting findings in text_
Being not from US or Mexico I was not familiar with the term 'Cinco de Mayo'. And finding it
among top 20 popular trigrams lead me to a very interesting reading on Wikipedia :)  


Another one was observing similar mentionings of Amazon among popular four-grams.
Tracing back to original sentences showed that there are many instances of almost the same text mentioning blogs on wordpress with random-generated-looking names.  
See examples below:  

<span style="color:gray">_
"JBFUA Blog (jbfuablog.wordpress.com) is a participant in the Amazon Services LLC and Amazon EU Associates Programmes designed to provide a means for sites to earn advertising fees by advertising and linking to amazon.com, amazon.ca, amazon.co.uk, amazon.de, amazon.fr, amazon.it and amazon.es."_</span>  

<span style="color:gray">_
"XVCYL Blog (xvcylblog.wordpress.com) is a participant in the Amazon Services LLC and Amazon EU Associates Programmes designed to provide a means for sites to earn advertising fees by advertising and linking to amazon.com, amazon.ca, amazon.co.uk, amazon.de, amazon.fr, amazon.it and amazon.es."_</span>  

I visited couple of them - all empty, seem machine generated and differ just by title, 
which is still always related to selling electronics. Looks like someone's affiliate marketing exercises :)


## Plans for Prediction Algorithm

The final prediction algorithm will be based on N-gram language model. 
Several tables will be built, containing n-grams accompanied with their probability rates. Next word will be proposed according to this rates.  In terms of parameters I'm now thinking to go till 5-grams, but this will depend on final accuracy tests.  

The smoothing will also be implemented to handle unseen n-grams. I'm going to experiment with 
Good-Turing and Katz's back-of and see what's better.  

The final application will be done with `Shiny` package and hosted on shinyapps.io server.
As hosting will be done under free account there can be limitations in terms of available RAM and server 
performance. One of the solutions will be to use `data.table` as it was advised to be the most efficient to store the final tables.
The tables will be also reduced as much as possible truncating low-frequency n-grams, it should still allow to keep up to 90% of word instances.  

Thanks for reading. Stay tuned! ;)